---
title: "relation between respitory diseases and air pollution"
author: "Jesper Smit, 2894491"
date: "`r Sys.Date()`"
output: pdf_document
bibliography: references.bib
csl: apa.csl

---

# Set-up your environment

```{r package_install, include=FALSE}

```

```{r packages}

library(tidyverse)
library(yaml)
library(rmarkdown)
library(janitor)
library(naniar)
library(tinytex)
library(ggplot2)

require(tidyverse)
require(rmarkdown)
require(yaml)
library(tidyr)
library(writexl)
library(readxl)
```

# Title Page
A new sentence
Include your names

Include the tutorial group number

Include your tutorial lecturer's name

# Part 1 - Identify a Social problem

## 1.1 Describe the Social Problem

Air pollution is the biggest current environmental health risk in the European Union, according to the European Environment Agency. The main risk of air pollution is respiratory disease, diseases like asthma or COPD. Each year, over 8 million people die from respiratory diseases (UT Health-San Antonio and South Texas Veterans Health Care System, 2021). In this paper, the relation between the severity of the air pollution and the frequency of respiratory disease is researched. 

# Part 2 - Data Sourcing

## 2.1 Load in the data

Preferably from a URL, but if not, make sure to download the data and store it in a shared location that you can load the data in from. Do not store the data in a folder you include in the Github repository!

```{r loading_data}
#loading the database 
ghs <- read_csv("ghs.csv")
gaq <- read_csv("gaq.csv")

```

midwest is an example dataset included in the tidyverse package

## 2.2 Provide a short summary of the dataset(s)


```{r}
#head(dataset)
```

In this case we see 28 variables, but we miss some information on what units they are in. We also don't know anything about the year/moment in which this data has been captured.

``` r
inline_code = TRUE
```

These are things that are usually included in the metadata of the dataset. For your project, you need to provide us with the information from your metadata that we need to understand your dataset of choice.

## 2.3 Describe the type of variables included

Think of things like:

-   Do the variables contain health information or SES information?

-   Have they been measured by interviewing individuals or is the data coming from administrative sources?

*For the sake of this example, I will continue with the assignment...*

# Part 3 - Quantifying

## 3.1 Data cleaning

Say we want to include only larger distances (above 2) in our dataset, we can filter for this.

```{r data_cleaning}
ghs <- janitor::clean_names(ghs)
gaq <- janitor::clean_names(gaq)

#voor de goeie data types per colum voor ghs en gaq
ghs <-
  ghs |>
  mutate(country = as_factor(country),
         gender = as_factor(gender),
         disease_name = as_factor(disease_name),
         disease_category  = as_factor(disease_category),
         age_group  = as_factor(age_group),
         treatment_type  = as_factor(treatment_type),
         availability_of_vaccines_treatment  = as_factor(availability_of_vaccines_treatment)
  )
gaq <-
  gaq |>
  mutate(country = as_factor(country)
  )

#alleen de rijen met Respiratory als disease_category voor ghs
ghs <- ghs %>%
  filter(disease_category == "Respiratory")
#alleen rijen van 2014 of later voor ghs
ghs <- ghs %>%
  filter(year >= 2014)
#alleen rijen tot 2025 
ghs <- ghs %>%
  filter(year >= 2014)
#verwijderd rijen met landen die niet in beide voorkomen
ghs <- ghs %>%
  filter(!country %in% c("Italy", "Canada", "Argentina"))
gaq <- gaq %>%
  filter(!country %in% c("Bangladesh", "Egypt", "Ethiopia", "Pakistan", "Phillipines", "Thailand", "Vietnam"))

gaq <- gaq %>% select(-state)


# gegevens van landen bij elkaar zetten dus geen states/provincies meer van global air pollution
gaq_country_year <- gaq %>%
  mutate(year = format(date, "%Y")) %>%
  group_by(country, year) %>%
  summarise(
    across(where(is.numeric), ~ mean(.x, na.rm = TRUE)),
    .groups = "drop"
  )
#onnodige kolommen verwijderen van global air pollution
gaq_country_year <- gaq_country_year %>%
  select(-c(wind_speed_km_h, humidity_percent, deforestation_rate_percent, aqi))
#onnodige kolommen verwijderen van global health statistics 
ghs_country_year <- ghs %>%
  select(-c(prevalence_rate_percent, age_group, gender, healthcare_access_percent, doctors_per_1000, hospital_beds_per_1000, treatment_type, availability_of_vaccines_treatment, average_treatment_cost_usd, dal_ys, improvement_in_5_years_percent))

# verschillende respiratory diseases van een jaar bij elkaar brengen 
ghs_country_year <- ghs_country_year %>%
  group_by(country, year) %>%
  summarise(
    population_affected = sum(population_affected, na.rm = TRUE),
    incidence_rate_percent = mean(incidence_rate_percent, na.rm = TRUE),
    mortality_rate_percent = mean(mortality_rate_percent, na.rm = TRUE),
    recovery_rate_percent = mean(recovery_rate_percent, na.rm = TRUE),
    per_capita_income_usd = mean(per_capita_income_usd, na.rm = TRUE),
    education_index = mean(education_index, na.rm = TRUE),
    urbanization_rate_percent = mean(urbanization_rate_percent, na.rm = TRUE),
    .groups = "drop"
  )





#year goeie datatype make in bijde datasets 
gaq_country_year <- gaq_country_year %>%
  mutate(year = as.numeric(year))

ghs_country_year <- ghs_country_year %>%
  mutate(year = as.numeric(year))


#zorgen dat alle namen van de landen in de 2 datasets gelijk zijn 
ghs_country_year$country <- countrycode(
  ghs_country_year$country,
  origin = "country.name",
  destination = "country.name"
)

gaq_country_year$country <- countrycode(
  gaq_country_year$country,
  origin = "country.name",
  destination = "country.name"
)

# variabele 1 bij genererate necessary variables leggen we uit wat ie doet.
gaq_country_year <- gaq_country_year %>%
  mutate(
    total_air_pollution = rowMeans(scale(select(., pm2_5_ug_m3, pm10_ug_m3,
                                                no_ug_m3, no2_ug_m3,
                                                n_ox_ppb, nh3_ug_m3,
                                                co_mg_m3, so2_ug_m3,
                                                o3_ug_m3, benzene_ug_m3,
                                                toluene_ug_m3, xylene_ug_m3)),
                                   na.rm = TRUE)
  )





#alleen de kolommen selecteren die bij elkaar mergen voor gaq
gaq_clean <- gaq_country_year %>%
  select(country, year, total_air_pollution)
#de databases mergen 
result <- gaq_clean %>%
  inner_join(ghs_country_year, by = c("country", "year"))

#variabele 2 en 3 bij genererate necessary variables leggen we uit wat ze doen.
result <- result %>%
  mutate(
    disease_intensity = scale(population_affected),
    disease_per_pollution = population_affected / (abs(total_air_pollution) + 1)
  )

#safe databases
saveRDS(gaq_country_year, "gaq_country_year.rds")
saveRDS(ghs_country_year, "ghs_country_year.rds")
saveRDS(result, "result.rds")



```

Please use a separate 'R block' of code for each type of cleaning. So, e.g. one for missing values, a new one for removing unnecessary variables etc.

## 3.2 Generate necessary variables

Variable 1

variabele maak ik eerder al in data cleaning, omdat ik voor data cleaning functies gebruik waar die variabele inzit. dus hier leg ik uit wat ze doen bij data cleaning staat de code aangegven met # welke code het is.

variabele 1: total air pollution is een index door de verschillende vervuilinge te standariseren en dan opt te tellen. uiteindelijk komt er een gemiddelde uit die vergeleken is met de vervuiling van alle lande in de dataset 
Waarde >0 meer vervuild dan gemiddeld, waarde = 0 gemiddeld vervuild, waarde 0< mnder vervuild gemiddeld.

```{r gen_var1}
 

```

Variable 2 en 3
variabele maak ik eerder al in data cleaning, omdat ik voor data cleaning functies gebruik waar die variabele inzit. dus hier leg ik uit wat ze doen bij data cleaning staat de code aangegven met # welke code het is.  

variabele 2: Disease intesity laat zien dit jaar  het meer of minder getroffen door respiratoire ziektes dan gemiddeld, 1 veel hogere ziekte last, 0 gemiddeld, -1 veel lagere ziekte last. Het neemt het gemiddelde van alle landen en rijen 
variabele 3: Disease per population ratio die laat zien hoe de pollution is in vergelijking met aantal diseases laagwaarde=minder ziekte relatief aan pollution, hoge waarde = veel ziekte ondanks lage/medium pollution
```{r gen_var2}




```

## 3.3 Visualize temporal variation

```{r}
library(dplyr)
library(tidyr)
library(ggplot2)


```

## 3.4 Visualize spatial variation

library(dplyr)
library(rnaturalearth)
library(rnaturalearthdata)
library(sf)
library(plotly)

# Data
data <- global_air_quality_2014_2025
names(data) <- make.names(names(data))

pm25_country <- data %>%
  group_by(Country) %>%
  summarise(PM2.5 = mean(PM2.5..ug.m3., na.rm = TRUE))

# Wereldkaart
world <- ne_countries(scale = "medium", returnclass = "sf")

# Merge
map_data <- world %>%
  left_join(pm25_country, by = c("name" = "Country"))

# Alleen landen met data
map_data <- map_data %>% filter(!is.na(PM2.5))

# Centroid per land
centroids <- st_centroid(map_data)

coords <- st_coordinates(centroids)
centroids$lon <- coords[,1]
centroids$lat <- coords[,2]

# Plotly puntenkaart
fig <- plot_ly(
  data = centroids,
  lon = ~lon,
  lat = ~lat,
  type = "scattergeo",
  mode = "markers",
  marker = list(
    size = 6,
    color = ~PM2.5,
    colorscale = "Viridis",
    reversescale = FALSE,
    colorbar = list(title = "PM2.5")
  ),
  text = ~paste("Land:", name, "<br>PM2.5:", PM2.5),
  hoverinfo = "text"
) %>%
  layout(
    title = "Interactieve Wereldkaart PM2.5 (gemiddelde 2014–2025)",
    geo = list(
      projection = list(type = "natural earth"),
      showland = TRUE,
      landcolor = "rgb(230, 230, 230)",
      showcountries = TRUE,
      countrycolor = "rgb(200, 200, 200)"
    )
  )

fig
Here you provide a description of why the plot above is relevant to your specific social problem.

## 3.5 Visualize sub-population variation


```{r}
#box plot
ggplot(result, aes(x = reorder(country, disease_intensity, median),
                   y = disease_intensity)) +
  geom_boxplot() +
  coord_flip() +
  labs(
    title = "Disease intensity by country",
    x = "Country",
    y = "Disease intensity"
  )
```


The boxplot reveals a high difference in disease intensity across countries. Russia and South Africa have a high median (the black lines in the boxes) compared with other countries; this suggests a higher disease burden than in other countries. Variability differs across countries, with some having a more unstable disease intensity (as seen by the size of the box). There are some outliers suggesting extreme spikes in disease intensity ( indicated by the dots), which may be temporary shocks or faults in the measurements. The results show a high variability between countries. 

## 3.6 Event analysis

Analyze the relationship between two variables.

```{r analysis}

```

Here you provide a description of why the plot above is relevant to your specific social problem.

# Part 4 - Discussion

## 4.1 Discuss your findings

# Part 5 - Reproducibility

## 5.1 Github repository link

Provide the link to your PUBLIC repository here: ...

## 5.2 Reference list

European Environment Agency. (2024, January 25). Air pollution. Www.eea.europa.eu. https://www.eea.europa.eu/en/topics/in-depth/air-pollution
‌

Levine, S., Marciniuk, D., Aglan, A., Celedón, J., Fong, K., Horsburgh, R., Malhotra, A., Masekela, R., Mortimer, K., Redde, H., Rice, M., & Simonds, A. (n.d.). UT Health-San Antonio and South Texas Veterans Health Care System. https://firsnet.org/wp-content/uploads/2025/01/FIRS_Master_09202021.pdf





